---
title: Hypothesis testing
output: 
  html_document: default
---

In this script we will see the preliminary hypothesis testing on the effects of sex on face shape, as well as the allometric and non-allometric decomposition of sexual dimorphism.
Because this is a sample testing, we will take a random sample of 200 males and 200 females to ensure a fast computation of these tests.

## Preliminaries

We will load the necessary libraries and datasets

```{r libraries dataset, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(ggpubr)
library(Hotelling)
library(tidyr)
library(vegan)
library(HDMD)
library(car) #for Manova
setwd('..')
path <- getwd()
setwd(paste(path, "/Results/FacePCA", sep = ""))
coeffs   <- read.csv("coeffs.csv")
eigenvec <- read.csv("eigenvectors.csv", header=F)
eigenval <- read.csv("eigenvalues.csv", header=F)
means    <- read.csv("means.csv", header=F)
facets   <- read.csv("facets.csv", header=F)
```

## Data cleaning

For data cleaning, we will remove every individual that is neither Male or Female, and individuals with NAs in any variable (Height, Weight, and Age).

```{r data cleaning}
coeffs = filter(coeffs, Sex == 'Female' | Sex == 'Male')
coeffs = droplevels(coeffs)
coeffs = na.omit(coeffs)
dim(coeffs)
head(coeffs[,1:8])
coeffs200 <- coeffs[c(sample(rownames(coeffs[coeffs$Sex == "Female",]), 200), 
                    sample(rownames(coeffs[coeffs$Sex == "Male",]), 200) ) , ]
```

After cleaning, we end up with `r nrow(coeffs)` individuals

## Facial sexual dimorphism analysis

### Differences in location

I will first test for differences in location between males in females in the face PCA.
To do that I will use two different methods.
In both methods I will use the full set of PCs, and a subset of the first 10 PCs.
First, I will compare males and females location using Hotelling's T^2^ statistic, which is a multivariate t-test.
The downside is that we cannot control for other covariates, such as Age and BMI.
The results are virtually identical when using a normalized PC space (not shown here).

```{r hotelling}
set.seed(10)
fit <- hotelling.test(.~Sex, data = coeffs200[,c(2, 7:ncol(coeffs))])
fit

fit10 <- hotelling.test(.~Sex, data = coeffs200[,c(2, 7:16)])
fit10
```

Second, I will use the more involved PERMANOVA (Anderson 2001) using the mahalanobis distance between males and females.
The benefit of mahalanobis distance is that the within-group variance is taken into account, and not only the euclidean distance in space. 

```{r permanova}
fit <- adonis(coeffs200[ , 7:ncol(coeffs200)] ~ coeffs200$Sex + coeffs200$Age + coeffs200$BMI,
              method="mahalanobis")
fit

fit10 <- adonis(coeffs200[ , 7:16] ~ coeffs200$Sex + coeffs200$Age + coeffs200$BMI, 
                method="mahalanobis")
fit10
```

We can also use the mahalanobis distance between group means as a measure of effect size.
In this case there is nothing else to compare with, so we show the mahalanobis distance between males and females in the whole dataset. 

```{r effect size}
madist <- pairwise.mahalanobis(coeffs200[, 7:ncol(coeffs200)], coeffs200$Sex)
madist$distance
```

### Differences in dispersion

I will use Anderson (2006) method for testing differences in dispersion between sexes

```{r permdist}
dist  <- vegdist(coeffs200[, 7:16], method = "mahalanobis")
fit10 <- betadisper(dist, coeffs200$Sex)
anova(fit10)
```


Remember to test the null hypothesis that the dispersions are identical between groups, using Anderson 2006

## Allometric and non-allometric components

In this section I will decompose the total facial sexual dimorphims (FSD) into its allometric and non-allometric components.
To generate the total FSD vector, we will regress face shape on sex, using age and BMI as covariates. 

```{r Total FSD and consensus faces}
#Multivariate regression
formula <- as.matrix(coeffs200[,7:93]) ~ coeffs200$Sex + coeffs200$Age + coeffs200$BMI + 
                                         coeffs200$Height
head(model.matrix(formula))
model   <- lm(as.matrix(coeffs200[,7:93]) ~ coeffs200$Sex + coeffs200$Age + coeffs200$BMI + 
                                            coeffs200$Height)
summary(manova(model), test="Pillai")
Manova(model)

model10   <- lm(as.matrix(coeffs200[,7:16]) ~ coeffs200$Sex + coeffs200$Age + coeffs200$BMI + 
                                            coeffs200$Height)
summary(manova(model10), test="Pillai")
Manova(model10)

#Calculating the average position
Males   <- colMeans(coeffs200[coeffs200$Sex == 'Male', 7:ncol(coeffs200)])
Females <- colMeans(coeffs200[coeffs200$Sex == 'Female', 7:ncol(coeffs200)])
consensus <- as.data.frame(t(data.frame(Males, Females)))
remove(Males, Females)
```

We can see the position of the average male and female faces in the first PCs.
The male and female consensus faces are highlighted to show their position in relation to the whole data set.

```{r consensus plots, fig.height=10, fig.width=10}
plotmeans <- function(matrix, means, factor, x1, y1)
{
  p1 <- ggplot(matrix, aes_string(x = x1, y = y1)) +
            geom_point(aes(colour = factor), alpha=0.1) +
            geom_point(data = means, aes_string(x = x1, y = y1), color="red", size=3) +
            theme_pubr()
  p1 <- ggpar(p1, palette = "jco")
  return(p1)
}
p1 <- plotmeans(coeffs200[,7:ncol(coeffs)], consensus, coeffs200$Sex, "PC1", "PC2")
p2 <- plotmeans(coeffs200[,7:ncol(coeffs)], consensus, coeffs200$Sex, "PC3", "PC4")
p3 <- plotmeans(coeffs200[,7:ncol(coeffs)], consensus, coeffs200$Sex, "PC5", "PC6")
p4 <- plotmeans(coeffs200[,7:ncol(coeffs)], consensus, coeffs200$Sex, "PC7", "PC8")
p5 <- plotmeans(coeffs200[,7:ncol(coeffs)], consensus, coeffs200$Sex, "PC9", "PC10")
p6 <- plotmeans(coeffs200[,7:ncol(coeffs)], consensus, coeffs200$Sex, "PC11", "PC12")

ggarrange(p1, p2, p3, p4, p5, p6, 
          ncol=2, nrow=3, 
          common.legend = T, legend = "top")
```

To obtain the allometric component, we use the previous multivariate linear regression. 
To decompose the total FSD into its allometric and non-allometric components, we use the estimated effect of height on the face, scaled by the mean difference between females and males.

```{r fsd estimations}
consensus[3,] <- model$coefficients[2,]
consensus[4,] <- coef(model)[5,] * (mean(coeffs200[coeffs200$Sex == 'Male',4]) -
                                    mean(coeffs200[coeffs200$Sex == 'Female',4]))
consensus[5,] <- as.numeric(consensus[3,]) - consensus[4,]
row.names(consensus) <- c("Males", "Females", "TotalSD", "Allometric", "Non-allometric")
```

We can graphically see the magnitude and direction of the vectors.
Note that in the plots, the vectors are scaled by 3.

```{r vectors, fig.height=10, fig.width=10, message=FALSE}
source("PlotPCs.R")
p1 <- plotvect(coeffs200[,7:ncol(coeffs200)], rownames(consensus)[3:5], consensus[3:5,], "PC1", "PC2")
p2 <- plotvect(coeffs200[,7:ncol(coeffs200)], rownames(consensus)[3:5], consensus[3:5,], "PC3", "PC4")
p3 <- plotvect(coeffs200[,7:ncol(coeffs200)], rownames(consensus)[3:5], consensus[3:5,], "PC5", "PC6")
p4 <- plotvect(coeffs200[,7:ncol(coeffs200)], rownames(consensus)[3:5], consensus[3:5,], "PC7", "PC8")
p5 <- plotvect(coeffs200[,7:ncol(coeffs200)], rownames(consensus)[3:5], consensus[3:5,], "PC9", "PC10")
p6 <- plotvect(coeffs200[,7:ncol(coeffs200)], rownames(consensus)[3:5], consensus[3:5,], "PC11", "PC12")

ggarrange(p1, p2, p3, p4, p5, p6, 
          ncol=2, nrow=3, 
          common.legend = T, legend = "top")
```

We export the fitted model and the vectors to visualize them as face shapes, [here](2017-07-FaceEffects.html).

```{r export}
setwd('..')
path <- getwd()
setwd(paste(path, "/Results/FSD", sep = ""))
write.csv(consensus, "vectors.csv", quote = F)
save(fit, file = "fitmodel.RData")
```