---
output: 
  html_document: default
---

This script will create a cluster of HapMap samples to use for a PCA, because we'll use only HapMap individuals to create the PCA, and then project everyone else. 

**Important:** check if hapmap samples contain only unrelated individuals. According to the file, they do, but double check

# Preliminares

Loading the libraries

```{r libraries}
library(dplyr)
```

Loading the database

```{r datasets}
setwd('..')
path <- getwd()
setwd(paste(path, "/DataBases/Genotypes", sep = ""))
famfile <- read.table("160603_UnionGenos_hg19hg38_IDSexParentsUpdated.fam", sep = "", header = F)
```

Filtering HapMap samples

```{r filtering}
famfile <- famfile %>% filter(V1 == "ASW" | V1 == "CEU" | V1 == "CHB" | V1 == "CHD" | V1 == "GIH" |
                              V1 == "JPT" | V1 == "LWK" | V1 == "MEX" | V1 == "MKK" | V1 == "TSI" |
                              V1 == "YRI")
famfile <- droplevels(famfile)
```
 
Creating a new file to export with one cluster per line
 
```{r new dataset}
newdat <- cbind("HAPM" , t(paste(famfile$V1, famfile$V2, sep="_")))
setwd('..')
path <- getwd()
setwd(paste(path, "/Databases/Genotypes", sep = ""))
write.table(cbind(famfile[,1:2], famfile[,1]), file="HapCluster.txt", quote = F, row.names = F, col.names = F)
```
