---
output: 
  html_document:
    fig_height: 10
    fig_width: 10
title: Population comparisons
---

In this script we'll look at the difference in patterns of facial sexual dimorphism between the populations previously defined. 
We'll use multivariate statistics to estimate group differences in the face PCA. 
Also, we'll estimate the vectors of FSD, as well as the allometric and non-allometric components between populations. 
To do that, we'll compare the magnitude between vectors (degree of facial sexual dimorphism), and the direction (pattern of facial sexual dimorphism). 

## Preliminaries

Loading libraries

```{r libraries data, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(ggpubr)
library(vegan)
library(HDMD)
library(car)
source("Distances.R")
#Color palette
OkabeIto <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2")
```

```{r datasets, message=FALSE, warning=FALSE}
setwd('..')
path <- getwd()
setwd(paste(path, "/Results/MergedData", sep = ""))
load("MergedDat.RData")
```

## Hypohtesis testing

It is clear from the scatter plots that Asian and South Asian samples differ substantially in their sample size. 
This could affect our estimations for differences.
For example, when estimating location differences between the groups, it is not possible to ascertain that that difference is due to location or dispersion, which may be affected by unbalanced data sets. 
Therefore, we'll first remove Asian samples from our location difference tests.
We'll test whether group averages differ between groups, by using a 5x2 design using continent and Sex as grouping variables.
As shown, both continent and Sex (and their interaction) are significant predictors of face shape.

```{r MANOVAS}
fit <- lm(as.matrix(dplyr::select(MergedDat[[1]], starts_with("Face") ) ) ~ 
                 MergedDat[[1]]$Sex * MergedDat[[1]]$prediction)
Manova(fit, type = "III")

fit5 <- lm(as.matrix(dplyr::select(MergedDat[[1]], starts_with("Face") ) )[,1:5] ~ 
                 MergedDat[[1]]$Sex * MergedDat[[1]]$prediction)
Manova(fit5, type = "III")

fit <- lm(as.matrix(dplyr::select(MergedDat[[1]], starts_with("Face") ) ) ~ 
                MergedDat[[1]]$Sex * MergedDat[[1]]$prediction + MergedDat[[1]]$Height +
                MergedDat[[1]]$BMI + MergedDat[[1]]$Age)
Manova(fit, type = "III")

fit5 <- lm(as.matrix(dplyr::select(MergedDat[[1]], starts_with("Face") ) )[,1:5] ~ 
                 MergedDat[[1]]$Sex * MergedDat[[1]]$prediction + MergedDat[[1]]$Height +
                MergedDat[[1]]$BMI + MergedDat[[1]]$Age)
Manova(fit5, type = "III")
```

```{r permanova}
#mahamatrix <- vegdist(MergedDat[[2]], method = "mahalanobis")
#fit2 <- betadisper(mahamatrix, MergedDat[[1]]$continent)
#plot(fit2)
#permutest(fit2, pairwise = T)

#fit2 <- adonis(as.matrix(MergedDat[[2]]) ~ MergedDat[[1]]$Sex * MergedDat[[1]]$continent, permutations = 999, 
#               method = "mahalanobis")
#fit2
```

We can also compute pairwise mahalanobis distance as a measure of effect size.
This is preferred from an Euclidean distance, because the within-group variation is taken into account in the mahalanobis distance.
As shown in the dendogram, EAS is the least similar group overall.

```{r mahal continent}
madistcont <- pairwise.mahalanobis(dplyr::select(MergedDat[[1]], starts_with("Face") ), 
                                   MergedDat[[1]]$prediction)
dimnames(madistcont$distance) <- list(levels(MergedDat[[1]]$prediction),
                                      levels(MergedDat[[1]]$prediction))
madistcont$distance
plot(hclust(as.dist(madistcont$distance)))
```

We can use the full 5 X 2 design to evaluate the overall distance between continental group divided by males and females
As seen in the dendogram, African and Asian faces cluster together, while American and European females, and American and European males form another cluster. 
Interestingly, American and European females cluster closer to the African branch.
Also, the continental group with the least difference between sexes is America, while the other groups have similar values.

```{r mahalanobis dist}
madist <- pairwise.mahalanobis(dplyr::select(MergedDat[[1]], starts_with("Face")),
                               interaction(MergedDat[[1]]$prediction, MergedDat[[1]]$Sex))
dimnames(madist$distance) <- list(levels(interaction(MergedDat[[1]]$prediction, MergedDat[[1]]$Sex)),
                                  levels(interaction(MergedDat[[1]]$prediction, MergedDat[[1]]$Sex)))
madist$distance
plot(hclust(as.dist(madist$distance)))
```

## Vectors comparison

Another approach is to calculate the sexual dimorphism vectors for each population, and compare them.
We can compare the magnitude and angle between vectors to assess the degree and pattern of facial sexual dimorphism respectively.
For the pattern of SD (angle) we can use the cosine similarity measure to assess statistical significance of pairwise comparisons. 
We can use permutation (permuting continent labels) to assess statistical significance.
We can obtain the vector of total sexual dimorphism with the beta coefficients from a regression of face shape on sex for each population.
To obtain the allometric component, we run a multivariate linear regression of face shape against Sex, Height, and continent. 
In this case, we'll also estimate the Height by continent interaction, so as to compare the vectors between continental groups, if not, they would have the same vector.

```{r regression models}
matrix  <- dplyr::select(MergedDat[[1]], starts_with("Face"))
covs    <- MergedDat[[1]] %>% dplyr::select(c("Sex", "Age", "Height", "BMI", "prediction"))
covs[,c(2:4)] <- scale(covs[,c(2:4)])
regressnon <- cbind(covs, matrix) %>% group_by(prediction) %>%
  do(model = lm(as.matrix(dplyr::select(., starts_with("Face"))) ~ Sex + Height + BMI + Age
                , data = .)) %>% tidy(model)

regresstot <- cbind(covs, matrix) %>% group_by(prediction) %>%
  do(model = lm(as.matrix(dplyr::select(., starts_with("Face"))) ~ Sex + BMI + Age
                , data = .))
regresstot <- regresstot %>% tidy(model)
```


```{r SD decomposition}
##TOTAL SD##
extractRegress <- function(dat, filter, select){
  coefs <- dat %>% group_by(prediction) %>% filter(term == filter) %>%
    dplyr::select(select)
  final <- as.data.frame(matrix(0, ncol = 88, nrow = 5))
  colnames(final) <- c("Pop", colnames(dplyr::select(MergedDat[[1]], starts_with("Face") ) ) )
  final$Pop       <- levels(MergedDat[[1]]$prediction)
  for(i in 1:nrow(final)){
    t <- coefs %>% filter(prediction == final[i,1])
    final[i,2:88] <- c(t[,select])[[1]]
  }
  return(final)
}

sexcoefstot   <- extractRegress(regresstot, "SexMale", "estimate")
sexcoefstotse <- extractRegress(regresstot, "SexMale", "std.error")
sexcoefstotunit <- sexcoefstot
sexcoefstotunit[,2:87] <- t(apply(sexcoefstot[,2:87], 1, function(x) x/norm(x, type = "2")))

sexcoefsnon   <- extractRegress(regressnon, "SexMale", "estimate")
sexcoefsnonse <- extractRegress(regressnon, "SexMale", "std.error")

heightcoefs   <- extractRegress(regressnon, "Height", "estimate")
heightcoefsse <- extractRegress(regressnon, "Height", "std.error")
```

We can estimate the angles between the total and the allo and non allo

```{r}
for(i in 1:5){
  an <- getAngle(sexcoefstot[i,2:88], sexcoefsnon[i,2:88])
  print(an * 180 / pi)
}

for(i in 1:5){
  an <- getAngle(sexcoefstot[i,2:88], heightcoefs[i,2:88])
  print(an * 180 / pi)
}
```


We can see the percentage of the total and non allo SD that's represented in the first 5 PCS

```{r 5PCs}
for(i in 1:5){
  perc <- norm(sexcoefstot[i,2:5], type = "2")^2 * 100 / norm(sexcoefstot[i,2:88], type = "2")^2
  print(perc)
}

for(i in 1:5){
  perc <- norm(sexcoefsnon[i,2:5], type = "2")^2 * 100 / norm(sexcoefsnon[i,2:88], type = "2")^2
  print(perc)
}

for(i in 1:5){
  perc <- norm(heightcoefs[i,2:5], type = "2")^2 * 100 / norm(heightcoefs[i,2:88], type = "2")^2
  print(perc)
}
```


```{r vector plots}
plotsquare <- function(PCA, x1, y1, vectors, se, xlim = NULL, ylim = NULL){
  minmax  <- data.frame(ymin = vectors[,y1] - se[,y1], ymax = vectors[,y1] + se[,y1],
                        xmin = vectors[,x1] - se[,x1], xmax = vectors[,x1] + se[,x1])
  vectors <- cbind(vectors, minmax)
  #lims    <- data.frame(y = max(abs(max(minmax[,1:2])), abs(min(minmax[,1:2]))), 
  #                      x = max(abs(max(minmax[,3:4])), abs(min(minmax[,3:4]))) )
  #lims <- lims + 0.2
  xname <- gsub("^.*?_", "" , x1)
  yname <- gsub("^.*?_", "" , y1)
  p1   <- ggplot(PCA, aes_string(x = x1, y = y1)) +
    geom_point(data = vectors, aes_string(x = x1, y = y1, color = "Pop"),
               size = 3) +
    geom_errorbar(data = vectors, aes(ymin = ymin, ymax = ymax, color = Pop)) + 
    geom_errorbarh(data = vectors, aes(xmin = xmin, xmax = xmax, color = Pop)) + 
    geom_segment(data = vectors, 
                 aes_string(x = 0, y = 0,
                            xend = paste(x1, "*1", sep = ""), 
                            yend = paste(y1, "*1", sep = ""),
                            color = "Pop"),
                 alpha = 0.3, size = 2, lineend = "butt") +
    #ylim(-lims$y, lims$y) +
    #xlim(-lims$x, lims$x) + 
    labs(x = xname, y = yname) +
    theme(legend.title = element_text(size=18)) + #Change legend size 
    theme(legend.text  = element_text(size=18)) + #Change text size
    guides(color = guide_legend(override.aes = list(size = 10))) + #Change legend size
    theme_pubr()
  
  if(is.null(xlim)){
  } else {
    p1 <- p1 + xlim(xlim[1], xlim[2])
  }
  if(is.null(ylim)){
  } else {
    p1 <- p1 + ylim(ylim[1], ylim[2])
  }
  
  p1 <- ggpar(p1, palette = OkabeIto)
  return(p1)
}

p1 <- plotsquare(MergedDat[[1]], "Face_PC1", "Face_PC2", sexcoefstot, sexcoefstotse, 
                 xlim = c(-0.1,2.8), ylim = c(-0.6, 1.8))
p2 <- plotsquare(MergedDat[[1]], "Face_PC3", "Face_PC4", sexcoefstot, sexcoefstotse, 
                 xlim = c(-0.5, 0.9), ylim = c(-0.5, 0.7))
p3 <- plotsquare(MergedDat[[1]], "Face_PC1", "Face_PC2", sexcoefsnon, sexcoefsnonse, 
                 xlim = c(-0.1,2.8), ylim = c(-0.6, 1.8))
p4 <- plotsquare(MergedDat[[1]], "Face_PC3", "Face_PC4", sexcoefsnon, sexcoefsnonse, 
                 xlim = c(-0.5, 0.9), ylim = c(-0.5, 0.7))
p5 <- plotsquare(MergedDat[[1]], "Face_PC1", "Face_PC2", heightcoefs, heightcoefsse, 
                 xlim = c(-0.1,2.8), ylim = c(-0.6, 1.8))
p6 <- plotsquare(MergedDat[[1]], "Face_PC3", "Face_PC4", heightcoefs, heightcoefsse, 
                 xlim = c(-0.5, 0.9), ylim = c(-0.5, 0.7))


ggarrange(p1, p2, p3, p4, p5, p6,
          nrow = 3, ncol = 2,
          common.legend = TRUE, legend = "top", 
          align = "v")
```


Angles between total SD vectors

```{r angles, message=F}
labscont   <- levels(MergedDat[[1]]$prediction)

totangle   <- getPAngle(sexcoefstot[,2:88])
totanglesub <- getPAngle(sexcoefstot[,2:5])
plot(hclust(as.dist(totangle)), labels = labscont, main = "Cluster Dendrogram Total SD")
plot(hclust(as.dist(totanglesub)), labels = labscont, main = "Cluster Dendrogram Total SD, subspace")

alloang    <- getPAngle(heightcoefs[,2:88])
alloangsub <- getPAngle(heightcoefs[,2:5])
plot(hclust(as.dist(alloang)), labels = labscont, main = "Cluster Dendrogram Allometric SD")
plot(hclust(as.dist(alloangsub)), labels = labscont, main = "Cluster Dendrogram Allometric SD, subspace")

nalloang    <- getPAngle(sexcoefsnon[,2:88])
nalloangsub <- getPAngle(sexcoefsnon[,2:5])
plot(hclust(as.dist(nalloang)), labels = labscont)
plot(hclust(as.dist(nalloangsub)), labels = labscont)

matrix  <- dplyr::select(MergedDat[[1]], starts_with("Face"))
covs    <- MergedDat[[1]] %>% dplyr::select(c("Sex", "Age", "Height", "BMI", "prediction"))
covs[,c(2:4)] <- scale(covs[,c(2:4)])

null <- getAngleNull(matrix,  
             covs[,1:4],
             covs$prediction,
             perm = 5000)

nulltot <- getAngleNull(matrix,  
             covs[,1:4],
             covs$prediction,
             perm = 5000,
             type = "total")

nullheight <- getAngleNull(matrix,  
             covs[,1:4],
             covs$prediction,
             perm = 100,
             maincov = "Height")

for(i in 1:10){
  pval <- sum(nulltot[i,] > totangle[lower.tri(totangle)][i]) / 5000
  print(c(totangle[lower.tri(totangle)][i], pval) )
}

for(i in 1:10){
  pval <- sum(null[i,] > nalloang[lower.tri(nalloang)][i]) / 5000
  print(c(nalloang[lower.tri(nalloang)][i], pval) )
}

for(i in 1:10){
  pval <- sum(nullheight[i,] > alloang[lower.tri(alloang)][i]) / 100
  print(c(alloang[lower.tri(alloang)][i], pval) )
}

#Using the subspace
null <- getAngleNull(matrix[,1:5],  
             covs[,1:4],
             covs$prediction,
             perm = 5000)

nulltot <- getAngleNull(matrix[,1:5],  
             covs[,1:4],
             covs$prediction,
             perm = 5000,
             type = "total")

for(i in 1:10){
  pval <- sum(nulltot[i,] > totanglesub[lower.tri(totanglesub)][i]) / 5000
  print(c(totanglesub[lower.tri(totanglesub)][i], pval) )
}

for(i in 1:10){
  pval <- sum(null[i,] > nalloangsub[lower.tri(nalloangsub)][i]) / 5000
  print(c(nalloangsub[lower.tri(nalloangsub)][i], pval) )
}
```

To export the PCA vectors and visually compare the effects of sex on facial morphology, we will standardize them so to have the same magnitude across populations.
To do that, we'll tranform each vector to a unit vector, using $\hat{u} = \frac{\overrightarrow{ {u} } {\|\overrightarrow{u}\|}$ 

```{r standardize magnitude, eval=FALSE}
popvectors <- rbind(totalSD, alloSD, nonalloSD)
cols       <- paste(c(rep("Total", 6), rep("Allo", 6), rep("NonAllo", 6)), popvectors[,1], sep = ".")
popvectors <- as.data.frame(t(popvectors[, (-1)]))
colnames(popvectors) <- cols

norm_vec <- function(x) sqrt(sum(x^2))
apply(popvectors, 2, function(x) norm_vec(x))

popvectunit <- as.data.frame(apply(popvectors, 2, function(x) x / norm_vec(x) ) )
apply(popvectunit, 2, function(x) norm_vec(x))
```

## Saving

```{r save, eval=FALSE}
setwd(paste(path, "/Results/FSD", sep = ""))
write.csv(popvectunit, file="popvectors.csv", quote = F, row.names = F)
```