---
title: Descriptive statistics
output:
  html_document: default
---

In this section we will look at descriptive statistics from covariates, and from the face PCA.

## Preliminaries

Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(GGally)
library(ggpubr)
library(ggridges)
library(ggrepel)
library(pophelper)
source("PlotFaces.R")
source("Distances.R")
source("PlotPCs.R")
```

Databases

```{r databases, warning=FALSE, message=FALSE}
setwd('..')
path <- getwd()
setwd(paste(path, "/Results/MergedData", sep = ""))
load("MergedDat.RData")
MergedDat[[1]]$Sex <- as.factor(MergedDat[[1]]$Sex)
MergedDat[[1]]$CS  <- log(MergedDat[[1]]$CS)

setwd(paste(path, "/Results/FacePCA", sep = ""))
face.eigenvalues  <- read_delim("eigenvalues.csv", delim = " ", col_names = FALSE)
face.eigenvectors <- read_csv("eigenvectors.csv", col_names = FALSE)
face.facets       <- read_csv("facets.csv", col_names = FALSE)
face.means        <- read_csv("means.csv", col_names = FALSE)

setwd(paste(path, "/Results/GenPCA", sep = ""))
gen.eigenvalues <- read_delim("PCA.eigenval", delim = " ", col_names = FALSE)
```

## Descriptive and exploratory analyses 

We can look at some summaries.
We have a total of `r nrow(MergedDat[[1]])` subjects with Genetic and Face data.

```{r summaries}
MergedDat[[1]] %>% dplyr::select(c("Age", "Height", "CS", "Weight", "BMI") ) %>%
  summarise_all(funs( mean(. , na.rm = T), sd(. , na.rm = T) )) %>% print()

MergedDat[[1]] %>% group_by(prediction, Sex) %>% dplyr::select(Age, Height, CS, Weight) %>% 
            summarise_all(funs(mean(. , na.rm = T), sd(. , na.rm = T))) %>% print()

sexbypop <- function(dat, var, y){
  p1 <- ggplot(dat, aes(interaction(Sex, prediction), var, colour = prediction)) +
    geom_boxplot() + 
    theme_pubr() + 
    labs(x = "Sex by population", colour = "Population", y = y) + 
    scale_x_discrete(labels=c())
  p1 <- ggpar(p1, palette = "jama")
  return(p1)
}

p1 <- sexbypop(MergedDat[[1]], MergedDat[[1]]$Height, "Height")
p2 <- sexbypop(MergedDat[[1]], MergedDat[[1]]$Age, "Age")
p3 <- sexbypop(MergedDat[[1]], MergedDat[[1]]$Weight, "Weight")
p4 <- sexbypop(MergedDat[[1]], MergedDat[[1]]$BMI, "BMI")
p5 <- sexbypop(MergedDat[[1]], MergedDat[[1]]$CS, "CS")

ggarrange(p1, p2, p3, p4, p5,
          common.legend = TRUE, legend = "right")
```

We can also explore some correlograms

```{r correlogram, fig.height = 8, fig.width = 12, fig.align = "center", message=FALSE, warning=FALSE}
p1 <- ggpairs(MergedDat[[1]], columns = c(3:8,10,11,60,61), ggplot2::aes(colour=Sex, alpha = 0.1), 
              lower = list(continuous = "smooth")) + theme_pubr()
p1 <- ggpar(p1, palette = "jama")
p1
```

### Face PCA

PCA screeploy

```{r face screeplot}
barplot(c(as.matrix(face.eigenvalues)), 
       main = "Face PCA Variances",
       xlab = "Principal Components",
       ylab = "Eigenvalue",
       col = "steelblue")
```

In the case of the Face PCA, the first 10 PCs account for `r sum(face.eigenvalues[1:10,])/(sum(face.eigenvalues) * 100 / 98) *100` percent of the total variance, while the first 4 PCs account for `r sum(face.eigenvalues[1:4,])/(sum(face.eigenvalues) * 100 / 98) *100`

We can use joy plots to compare the distributions of each face PC grouped by sex.
Note that I'll use the mahalanobis space instead of the euclidean one.

```{r face joyplots, fig.height = 10, fig.width = 8, fig.align = "center", message=FALSE}
ts <- MergedDat[[1]] %>% select(starts_with("Face"))
p1 <- PlotPCs(ts, MergedDat[[1]]$Sex ,"Face_PC1", "Face_PC9")
p2 <- PlotPCs(ts, MergedDat[[1]]$Sex, "Face_PC10", "Face_PC19")
p3 <- PlotPCs(ts, MergedDat[[1]]$Sex, "Face_PC20", "Face_PC29")
p4 <- PlotPCs(ts, MergedDat[[1]]$Sex, "Face_PC30", "Face_PC39")
remove(ts)
ggarrange(p1, p2, p3, p4,
          common.legend = T, legend = "bottom")
```

We can also see the distribution in the first PCs of sexes by each continental group

```{r sexes by continent plots, fig.height = 14, fig.width = 10}
plotsexes <- function(PCA, cont, x1, y1){
  highPCA <- filter(PCA, cont)
  backPCA <- filter(PCA, !cont)
  sexes   <- filter(MergedDat[[1]], cont)$Sex
  title   <- as.character(MergedDat[[1]]$prediction[cont][1])
  avgs    <- highPCA %>% select(x1, y1, Sex) %>% 
                  group_by(Sex) %>% summarise_all(funs(mean))
  p1 <- ggplot(highPCA, aes_string(x = x1, y = y1)) + 
            geom_point(aes(color = sexes)) +
            geom_point(data = avgs, size = 5, aes(color = Sex)) +
            geom_point(data = avgs, size = 3, color = "black") +
            geom_point(data = backPCA, alpha = 0.01, color = "black", aes_string(x = x1, y = y1)) + 
            theme_pubr() + 
            ggtitle(title)
  p1 <- ggpar(p1, palette = "jco")
  return(p1)
}

p1 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "AFR", "Face_PC1", "Face_PC2")
p2 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "AFR", "Face_PC3", "Face_PC4")
p3 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "AMR", "Face_PC1", "Face_PC2")
p4 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "AMR", "Face_PC3", "Face_PC4")
p5 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "EAS", "Face_PC1", "Face_PC2")
p6 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "EAS", "Face_PC3", "Face_PC4")
p7 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "SAS", "Face_PC1", "Face_PC2")
p8 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "SAS", "Face_PC3", "Face_PC4")
p9 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "MDE", "Face_PC1", "Face_PC2")
p10 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "MDE", "Face_PC3", "Face_PC4")
p11 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "EUR", "Face_PC1", "Face_PC2")
p12 <- plotsexes(MergedDat[[1]], MergedDat[[1]]$prediction == "EUR", "Face_PC3", "Face_PC4")

ggarrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12,
          ncol = 2, nrow = 6,
          common.legend = TRUE, legend = "bottom", 
          align = "v")
```

### Genetic PCA

Similarly, we can look at the genetic PCs with respect to populations

PCA screeplot

```{r screeplot, fig.height=4, fig.width=8} 
barplot(c(as.matrix(gen.eigenvalues)), 
       main = "Genetic PCA Variances",
       xlab = "Principal Components",
       ylab = "Eigenvalue",
       col = "steelblue")
```

In the case of the Genetic PCA, the first 10 PCs account for no more than `r sum(gen.eigenvalues[1:10,])/sum(gen.eigenvalues) *100` percent of the total variance, while the first 4 PCs account for no more than `r sum(gen.eigenvalues[1:4,])/sum(gen.eigenvalues) *100`.

We can use joy plots to compare the distributions of each genetic PC grouped by population.

```{r joyplots, message=FALSE}
source("PlotPCs.R")
ts <- MergedDat[[1]] %>% select(starts_with("Gen"))
p1 <- PlotPCs(ts, MergedDat[[1]]$prediction, "PC1", "PC9")
p2 <- PlotPCs(ts, MergedDat[[1]]$prediction, "PC10", "PC19")
p3 <- PlotPCs(ts, MergedDat[[1]]$prediction, "PC20", "PC29")
p4 <- PlotPCs(ts, MergedDat[[1]]$prediction, "PC30", "PC39")
ggarrange(p1, p2, p3, p4,
          common.legend = TRUE, legend = "right", 
          align = "v")

```

This is the scatter plot of the 1000 Genomes populations.
These are the at the 1000Genomes populations: 

|POP | SUPER POP | Description                 
|----|-----------|------------------------------------------------------
|FIN | EUR       | Finnish in Finland 	
|CHS | EAS       | Han Chinese South
|GBR | EUR       | British from England and Scotland
|PUR | AMR       | Puerto Rican in Puerto Rico
|CLM | AMR       | Colombian in Medellin, Colombia
|MXL | AMR       | Mexican Ancestry in Los Angeles, California
|TSI | EUR       | Toscani in Italia
|LWK | AFR       | Luhya in Webuye, Kenya
|JPT | EAS       | Japanese in Tokyo, Japan
|IBS | EUR       | Iberian populations in Spain
|PEL | AMR       | Peruvian in Lima, Peru
|CDX | EAS       | Chinese Dai in Xishuangbanna, China
|YRI | AFR       | Yoruba in Ibadan, Nigeria 
|KHV | EAS       | Kinh in Ho Chi Minh City, Vietnam
|ASW | AFR       | African ancestry in Southwest USA
|ACB | AFR       | African Caribbean in Barbados
|CHB | EAS       | Han Chinese in Beijing, China
|GIH | SAS       | Gujarati Indians in Houston, Texas
|GWD | AFR       | Gambian in Western Division, The Gambia
|PJL | SAS       | Punjabi in Lahore,Pakistan
|MSL | AFR       | Mende in Sierra Leone
|BEB | SAS       | Bengali in Bangladesh
|ESN | AFR       | Esan in Nigeria
|STU | SAS       | Sri Lankan Tamil in the UK
|ITU | SAS       | Indian Telugu in the UK
|CEU | EUR       | Utah residents with Northern and Western European ancestry from the CEPH collection


```{r 1000genomes}
plotoneg <- function(dat, x1, y1){
  p1 <- ggplot(dat, aes_string(x = x1, y = y1, color = "super_pop")) +
            geom_point(alpha = 0.3, size = 2) + labs(color = "Pop") +
            theme_pubr()
  p1 <- ggpar(p1, palette = "jama")
  return(p1)
}

p1 <- plotoneg(MergedDat[[2]], "PC1", "PC2")
p2 <- plotoneg(MergedDat[[2]], "PC3", "PC4")
p3 <- plotoneg(MergedDat[[2]], "PC5", "PC6")
p4 <- plotoneg(MergedDat[[2]], "PC7", "PC8")
p5 <- plotoneg(MergedDat[[2]], "PC9", "PC10")
p6 <- plotoneg(MergedDat[[2]], "PC11", "PC12")

ggarrange(p1, p2, p3, p4, p5, p6, 
          nrow = 3, ncol = 2,
          common.legend = TRUE, legend = "right", 
          align = "v")
```

We can plot our samples with the centroid of the HapMap pops indicated.

```{r 1000g plus adapt, fig.width=10, fig.height=4}
meansone <- MergedDat[[2]] %>% group_by(pop, super_pop) %>% 
            select(contains("PC")) %>% summarise_all(funs(mean))

plotonesamp <- function(dat1, dat2, x1, y1){
  x2 <- paste("Gen_", x1, sep = "")
  y2 <- paste("Gen_", y1, sep = "")
  colnames(dat2)[which(colnames(dat2) == x2)] <- x1
  colnames(dat2)[which(colnames(dat2) == y2)] <- y1
  p1 <- ggplot(data = dat2, aes_string(x1, y1)) + 
                 geom_point(aes(colour = prediction), alpha = 0.3) +
                  ggrepel::geom_label_repel(data = dat1, aes(label = pop), 
                            alpha = 0.8) + 
          theme_pubr(legend = "none")
  p1 <- ggpar(p1, palette = "jama")
  return(p1)
}
  
p1 <- plotonesamp(meansone, MergedDat[[1]], "PC1", "PC2")
p2 <- plotonesamp(meansone, MergedDat[[1]], "PC3", "PC4")

ggarrange(p1, p2,
          nrow = 1, ncol = 2,
          align = "v")
```

Finally, let's take a look at the admixture analysis.
Based on visual inspection, these are the approximate cluster by geography associations

|Cluster | POP|
|--------|----|
|1       | NEU|
|2       | EAS|
|3       | SAS|
|4       | SEU|
|5       | AMR|
|6       | AFR|


```{r admixture, fig.width=10}
groups <- tibble(IID = rownames(MergedDat[[3]]$Merge1000Gsamples.6.Q))
groups <- groups %>% left_join(MergedDat[[1]], by = "IID") %>% select(IID, prediction)
groups <- groups %>% left_join(MergedDat[[2]], by = "IID") %>% select(IID, prediction, super_pop, pop)
groups <- groups %>%
  mutate(super_pop = ifelse(is.na(prediction), as.character(super_pop), as.character(prediction)),
         sample = ifelse(is.na(prediction), "1000Genomes", "ADAPT"),
         pop = ifelse(is.na(prediction), as.character(pop), as.character(prediction))) %>%
  select(IID, super_pop, pop, sample)

cleanadmix <- MergedDat[[3]]$Merge1000Gsamples.6.Q %>% filter(!is.na(groups$pop))
groups     <- groups %>% filter(!is.na(groups$pop))
cleanadmix <- list("sample1"=cleanadmix)

p1 <- plotQ(cleanadmix, returnplot=T, exportplot=F, quiet=T, basesize=11, showsp=F,
            grplab=as.data.frame(groups[,c(4,2,3)]), ordergrp=TRUE, grplabangle = 90)
plot(p1$plot[[1]])
```

## Face plots

Here we'll look at some basic 3D plots of average facial morphology looking at the differences between populations and sexes

```{r }
avgcoords    <- MergedDat[[1]] %>% group_by(prediction, Sex) %>% select(contains("Face_")) %>% 
                    summarise_all(funs(mean))
avglandmarks <- apply(as.matrix(avgcoords[, 3:89] * 1) %*% t(as.matrix(face.eigenvectors)), 1, 
                              function(x) x + t(face.means))

source("Distances.R")
eucdist <- tibble(AFR = rep(0, nrow(face.means)/3), 
                  AMR = rep(0, nrow(face.means)/3), 
                  EAS = rep(0, nrow(face.means)/3),
                  EUR = rep(0, nrow(face.means)/3),
                  SAS = rep(0, nrow(face.means)/3)) 
                      
col <- 1
range01 <- function(x){(x-min(x))/(max(x)-min(x))}
for(i in seq(1, 10, 2)){
  eucdist[, col] <- getDistance(avglandmarks[,i], avglandmarks[,i+1])
  eucdist[, col] <- range01(eucdist[, col])
  col <- col + 1
}

source("PlotFaces.R")

#PlotMultipleFaces(avglandmarks[,1], as.data.frame(face.facets), as.data.frame(eucdist))
PlotComparisonFaces(avglandmarks, as.data.frame(face.facets))
```



