---
title: Face effects
output: 
  html_document: default
---

In this script we will look at some interactive plots showing the effect of sex on the face. 
Because this is a preliminary assessment, the effect of sex is estimated using the whole sample (without taking their ancestries into account).
This script is built from the [hypothesis testing](2017-07-HTesting.html) in which I created the vectors of sexual dimorphism.
Remember to open the file in chrome.

## Preliminaries

We will load the necessary libraries, and datasets

```{r libraries, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
library(tidyr)
setwd('..')
path <- getwd()
setwd(paste(path, "/Results/FacePCA", sep = ""))
coeffs   <- read.csv("cleancoeffs.csv")
eigenvec <- read.csv("eigenvectors.csv", header=F)
eigenval <- read.csv("eigenvalues.csv", header=F)
means    <- read.csv("means.csv", header=F)
facets   <- read.csv("facets.csv", header=F)
setwd(paste(path, "/Results/FSD", sep = ""))
load("fitmodel.RData")
vectors <- read.csv("vectors.csv", row.names = 1)
```

## Face plots

In this section we will see the interactive plots showing facial sexual dimorphism.
To do that we need to transform the PCA vectors to the shape coordinates.
Each vector (total, allometric and non-allometric) is multiplied by a constant to exaggerate the facial effects (in this case 2 and -2), multiplied by the matrix of eigenvectors, and then added the means for each landmark.

```{r fsd coordinates}
fsdcoords <- data.frame(Tot.male = rep(0, nrow(means)), Tot.female = rep(0, nrow(means)), 
                        Allo.male = rep(0, nrow(means)), Allo.female = rep(0, nrow(means)),
                        Nallo.male = rep(0, nrow(means)), Nallo.female = rep(0, nrow(means)))
fsdcoords[,c(1,3,5)] <- apply(as.matrix(vectors[3:5,] * -2) %*% t(as.matrix(eigenvec)), 1, 
                              function(x) x + t(means))
fsdcoords[,c(2,4,6)] <- apply(as.matrix(vectors[3:5,] * 2) %*% t(as.matrix(eigenvec)), 1, 
                              function(x) x + t(means))
averagecoords <- t(( colMeans(vectors[1:2,]) %*% t(as.matrix(eigenvec))) + t(means))
```

The first set of plots show the overlaying of the two face shapes.
In this case male faces are shown as turquoise, and females as wheat color.

```{r 3d fsd plots}
source("PlotFaces.R")
Plot2Faces(fsdcoords[ ,1], fsdcoords[,2], facets, "Total SD")
Plot2Faces(fsdcoords[ ,3], fsdcoords[,4], facets, "Allometric SD")
Plot2Faces(fsdcoords[ ,5], fsdcoords[,6], facets, "Non allometric SD")
```

```{r distance calculation}
source("Distances.R")
eucdist <- data.frame(Totdist = rep(0, nrow(means)), 
                      Allodist =  rep(0, nrow(means)), 
                      Nallodist = rep(0, nrow(means)))

eucdist[,1] <- getDistance(fsdcoords[ ,1], fsdcoords[,2])
eucdist[,2] <- getDistance(fsdcoords[ ,3], fsdcoords[,4])
eucdist[,3] <- getDistance(fsdcoords[ ,5], fsdcoords[,6])

#On the same scale
eucdist <- gather(eucdist, type, dist)
eucdist$dist <- scales::rescale(eucdist$dist)
```

In this other set of plots you can see the euclidean distance between the two morphs, as a colormap in the average face.
At this point, the colormap is not scaled to be comparable across plost, and the legend does not show correctly the euclidean distances.

```{r distplots}
source("PlotFaces.R")
Plot1Face(averagecoords, facets, colormap = c(as.matrix(eucdist %>% filter(type == "Totdist") %>% 
                                                        dplyr::select(dist))), title = "Total SD")
Plot1Face(averagecoords, facets, colormap = c(as.matrix(eucdist %>% filter(type == "Allodist") %>% 
                                                        dplyr::select(dist))), title = "Allometric SD")
Plot1Face(averagecoords, facets, colormap = c(as.matrix(eucdist %>% filter(type == "Nallodist") %>% 
                                                    dplyr::select(dist))), title = "Non allometric SD")
```
